{\bf Intrusion Detection using Parzen-Windows on Provenance Graph Statistics}

### 2013-12-10 writeup/paper-final.tex
little issues

\begin{abstract}
Provenance is data that describes how a digital artifact came to be in its current state. We hypothesize
that intrusions on a system leave behind anomalies in the lineage of digital artifacts.
We present an intrusion detection approach to find these anomalies by analyzing centrality metrics
on provenance graphs. We use a Parzen-Window approach \cite{parzen} on various provenance graph centrality metrics \cite{clustering}
to determine probability density estimates of normal behavior, and we use these density estimates to 
determine if an intrusion occurred. We used this approach to analyze {\em user-to-remote} (u2r) intrusions 
from the 1998 DARPA Intrusion Detection data sets \cite{darpa} and 
achieved up to *TODO true positive rate for intrusions* accuracy in detecting
intrusions with only *TODO false positive rate for intrusions* accuracy. We also present future work
to extend our intrusion model to an online intrusion detection system.

@@

Provenance is data that describes how digital artifacts
came to be in their current state. Provenance data is typically structured as a directed acyclic graph with
typed nodes and typed edges.
Nodes typically include processes and files, and edges are directed from nodes to their dependencies. For example,
process nodes have edges directed towards input file nodes and to the parent process's node, and
file nodes have edges directed towards previous versions of the file and to the nodes of the
processes that modified the file.

@@

>>> TODO INSERT PROVENANCE GRAPH SMALL EXAMPLE HERE SHOWING DATA FLOW

@@

POSSIBLY INCLUDE THIS
* the papers suck at finding u2r
* because they don't look at provenance graph structure
* we can do better?

@@

@@

@@

Given progressive forms of these statistics, we can modify PASS to compute Parzen-Window density estimates as it generates provenance, or we can audit system calls like SPADE \cite{spade} to build provenance graphs in user space and apply our Parzen-Window technique.
\subsection{Local vs. Global}
In our experiments, we looked at provenance graphs in their entirety. However, provenance graphs are meant to be long-lived, and an intrusion that affected only a small portion of the graph will have negligible effects on the global structure of the graph. To resolve this issue, a possible future extension to our work is to cluster the nodes based on time \cite{clustering} and then only examine the most ``recent" nodes when applying our histogram technique. Another approach would be to discard provenance data in computing Parzen-Window density estimates after the data have reached a certain age.

>>>>

\begin{abstract}
Security rests on being able to identify potentially malicious behavior.
One way to identify malicious behavior is to compare system behavior to known normal behavior.
Intrusions on a system leave typically leave behind digital artifacts, and we hypothesize
that we can detect intrusions on a system by identifying anomalies the artifacts' provenance, the lineage of how
the artifact came to be in its present state.
We present an intrusion detection approach to find these anomalies by analyzing centrality metrics
on provenance graphs. We use a Parzen-Window approach \cite{parzen} on various provenance graph centrality metrics \cite{clustering}
to determine probability density estimates of normal behavior, and we use these density estimates to 
determine if an intrusion occurred. We used this approach to analyze {\em user-to-remote} (u2r) intrusions 
from the 1998 DARPA Intrusion Detection data sets \cite{darpa} and 
achieved up to *TODO true positive rate for intrusions* accuracy in detecting
intrusions with only *TODO false positive rate for intrusions* accuracy. We also present future work
to extend our intrusion model to an online intrusion detection system.

@@

Provenance is data that describes how digital artifacts
came to be in their current state. Provenance data is typically structured as a directed acyclic graph with
typed nodes and typed edges.
Nodes typically include processes and files, and edges are directed from nodes to their dependencies. For example,
process nodes have edges directed towards input file nodes and to the parent process's node, and
file nodes have edges directed towards previous versions of the file and to the nodes of the
processes that modified the file. See Figure \ref{graph-simple} for an illustration of these provenance
graph node and edge types.

@@

\begin{figure}
  \label{graph-simple}
  \centering
    \includegraphics[width=0.5\textwidth]{img/graph-simple.png}
    \caption{Example of a provenance graph, showing the types of nodes and edges common in provenance graphs} 
\end{figure}

@@

POSSIBLY INCLUDE THIS TODO
* the papers suck at finding u2r
* because they don't look at provenance graph structure
* we can do better?

@@

We envision a provenance aware system which constantly builds normal profiles of processes. Simultaneously,
the system looks for instances of deviation from normal behavior and handles them appropriately
(e.g., terminating the process, slowing down the process). In our work, we derive normal profiles of process behavior
from provenance graphs, and we use a Parzen-Window model on various provenance graph statistics
to determine if anomalous nodes appear in the graph.

@@

\subsection{Limitations}

@@

Given progressive forms of these statistics, we can modify PASS to compute Parzen-Window density estimates as it generates provenance, or we can audit system calls like SPADE \cite{spade} to build provenance graphs in user space and apply our Parzen-Window technique.
\subsection{Local vs. Global}
In our experiments, we looked at provenance graphs in their entirety. However, provenance graphs are meant to be long-lived, and an intrusion that affected only a small portion of the graph will have negligible effects on the global structure of the graph. To resolve this issue, a possible future extension to our work is to cluster the nodes based on time \cite{clustering} and then only examine the most ``recent" nodes when applying our histogram technique. Another approach would be to discard provenance data in computing Parzen-Window density estimates after the data have reached a certain age. By allowing the model to evolve over time, the system
can combat an {\em autoimmune disorders} in which attackers purposely invoke programs in weird ways to fool the system into thinking the programs are
behaving abnormally.
